How Tree-like are the data?

### 2019-04-30 phyloglmm_ms.tex
minor change

Ever-increasing data collection capabilities (e.g. genomic sequencing, telemetry studies of animal behaviour, or environmental remote sensing), in combination with large-scale synthetic data bases of species occurrence and phenotypic traits, are making larger volumes of biological data available over an ever-wider taxonomic range.
Researchers use these data to fit complex models describing species occurrence and traits.
In particular, phylogenetic comparative analysis (PCA) using phylogenetic (eco-evolutionary) regression \citep{hansen2012interpreting} is becoming increasingly popular. % \bmb{might be interesting to do some bibliometry, although absolutely not necessary --- i.e. how do numbers of citations of (whatever) increase relative to the overall size of the eco/evo literature? \ldots}
PCAs explore the relationships among species traits or distributions while taking the underlying evolutionary relationships of the species into account; they can be used to control statistically for phylogenetic relationships, to quantify the phylogenetic signals (a measure of dependence among species responses due to their phylogenetic relationships) in trait distributions, or both.
Unlike standard statistical models, where all of the predictor variables of interest (for example, species traits and environmental factors) are directly observable,  PCAs use phylogenetic relationships to estimate the unobserved process of trait evolution \citep{felsenstein1985phylogenies, butler2004phylogenetic}. 
% While evolutionary relationships are biologically important and interesting, they often get neglected in applications and multi-species analyses \citep{bunnefeld2012island, santamaria2012evolution}. 
Ecologists have used phylogenetic relationships in multi-species models \citep{garland1992procedures, freckleton2002phylogenetic, ord2010adaptation, davies2013phylogenetic}; more recently, they have begun to integrate evolutionary knowledge in applied ecological studies addressing biodiversity conservation and the effects of climate change \citep{winter2013phylogenetic, santamaria2012evolution, lankau2011incorporating, lavergne2010biodiversity, mace2008evolutionary}.
While a wide range of tools is available for comparative analyses, existing procedures may be either insufficiently flexible or too computationally demanding when analyzing large volumes of data.
In such cases, researchers must find ways to simplify their analyses: for example, treating species effects as independent, thus neglecting phylogenetic correlations among species responses \citep{bunnefeld2012island}; ignoring degrees of relatedness and treating taxon as a strictly hierarchical description \citep{tella1999habitat}; or neglecting within-species variation \citep{ord2010adaptation}.
%% bunnefeld2012island: used taxon for random slopes but did not include phylogenetic information
%% ord2010adaptation : used OU process with means (single obs per taxa),
%%  i.e. ignoring within-taxon variation?
%% fix me!?
% which can lead to statistical issues and erroneous conclusions
% \citep{felsenstein1985phylogenies, li2017statistical}.
% \bmb{What precise ``issues'' do you have in mind? Don't think we need to state them here, but we should know. Based on your comments, I can guess that (1) by using random slopes across taxa, Bunnefeld 2012 is neglecting more subtle variation due to differential relatedness? (2) not sure whether taking taxon-level means (Ord 2010) is actually a problem. Ignores some variation,
% but it might be ignorable (cf. Murtaugh 2007 ``Simplicity and complexity in ecological data analysis'')}
In this paper, we propose an alternative method for flexibly and efficiently modeling phylogenetic relationships via mixed effect models through random effects. 
This method allow researchers to easily incorporate various complexities without sacrificing speed.

>>>>

Ever-increasing data collection capabilities (e.g. genomic sequencing, telemetry studies of animal behaviour, or environmental remote sensing), in combination with large-scale synthetic data bases of species occurrence and phenotypic traits, are making larger volumes of biological data available over an ever-wider taxonomic range.
Researchers use these data to fit complex models describing species occurrence and traits.
In particular, phylogenetic comparative analysis (PCA) using phylogenetic (eco-evolutionary) regression \citep{hansen2012interpreting} is a powerful technique to explore relationships among species traits or distributions.
% \bmb{might be interesting to do some bibliometry, although absolutely not necessary --- i.e. how do numbers of citations of (whatever) increase relative to the overall size of the eco/evo literature? \ldots}
Given a known phylogenetic tree, PCAs explore the relationships among species traits or distributions while taking the underlying evolutionary relationships of the species into account; they can be used to control statistically for phylogenetic relationships, to quantify the phylogenetic signals (a measure of dependence among species responses due to their phylogenetic relationships) in trait distributions, or both.
Unlike standard statistical models, where all of the predictor variables of interest (for example, species traits and environmental factors) are directly observable,  PCAs use phylogenetic relationships to estimate the amount of unobserved process of trait evolution (i.e how "tree-like" are the observed data given the phylogenetic tree) \citep{felsenstein1985phylogenies, butler2004phylogenetic}. 
% While evolutionary relationships are biologically important and interesting, they often get neglected in applications and multi-species analyses \citep{bunnefeld2012island, santamaria2012evolution}. 
Ecologists have used phylogenetic relationships in multi-species models \citep{garland1992procedures, freckleton2002phylogenetic, ord2010adaptation, davies2013phylogenetic}; and more recently becoming increasingly popular, they have begun to integrate evolutionary knowledge in applied ecological studies addressing biodiversity conservation and the effects of climate change \citep{winter2013phylogenetic, santamaria2012evolution, lankau2011incorporating, lavergne2010biodiversity, mace2008evolutionary}.
While a wide range of tools is available for comparative analyses, existing procedures may be either insufficiently flexible or too computationally demanding when analyzing large volumes of data.
In such cases, researchers must find ways to simplify their analyses: for example, treating species effects as independent, thus neglecting phylogenetic correlations among species responses \citep{bunnefeld2012island}; ignoring degrees of relatedness and treating taxon as a strictly hierarchical description \citep{tella1999habitat}; or neglecting within-species variation \citep{ord2010adaptation}.
%% bunnefeld2012island: used taxon for random slopes but did not include phylogenetic information
%% ord2010adaptation : used OU process with means (single obs per taxa),
%%  i.e. ignoring within-taxon variation?
%% fix me!?
% which can lead to statistical issues and erroneous conclusions
% \citep{felsenstein1985phylogenies, li2017statistical}.
% \bmb{What precise ``issues'' do you have in mind? Don't think we need to state them here, but we should know. Based on your comments, I can guess that (1) by using random slopes across taxa, Bunnefeld 2012 is neglecting more subtle variation due to differential relatedness? (2) not sure whether taking taxon-level means (Ord 2010) is actually a problem. Ignores some variation,
% but it might be ignorable (cf. Murtaugh 2007 ``Simplicity and complexity in ecological data analysis'')}
In this paper, we propose an alternative method for flexibly and efficiently modeling phylogenetic relationships via mixed effect models through random effects. 
This method allow researchers to easily incorporate various complexities without sacrificing speed.

### 2019-04-30 phylotreelike.rmd
minor change

```{r, sampling, echo=FALSE,cache=TRUE}
ddprop <- data.frame(nspp=3:40, prop=NA)
rep_pp <- pp
for(i in 3:40){
phy <- rtree(n = i)
Vphy <- vcv(phy)

Y_phy <- MASS::mvrnorm(n=10000
  , mu=rep(0,each=i)
  , Sigma=Vphy
)

@@

c <- 1.96

dd2 <- (dd
  %>% rowwise()
  %>% mutate(LR = ll-null
             # , LRT = max(as.numeric(LR > log(c)),0.5)
             , LRT = LR > log(c)
             )
  # %>% arrange(LR)
)

@@

c <- 1.96

dd2 <- (dd
  %>% mutate(LR = ll-null # line log ratio
       , sumLL = cumsum(ll)
       , sumnull = cumsum(null)
  )
  %>% rowwise()
  %>% mutate(LRT = LR > log(c)
      , sumLR = sumLL - sumnull
      , sumLRT = sumLR > log(c)
      )
  # %>% arrange(LR)
)

ddsort <- (dd
  %>% mutate(LR = ll-null) # line log ratio
  %>% arrange(LR)
  %>% mutate(sumLL = cumsum(ll)
       , sumnull = cumsum(null)
  )
  %>% rowwise()
  %>% mutate(LRT = LR > log(c)
      , sumLR = sumLL - sumnull
      , sumLRT = sumLR > log(c)
      )
  # %>% arrange(LR)
)

>>>>

```{r, sampling, echo=FALSE,cache=TRUE}
ddprop <- data.frame(nspp=3:100, prop=NA)
rep_pp <- pp
for(i in 3:100){
phy <- rtree(n = i)
Vphy <- vcv(phy)

Y_phy <- MASS::mvrnorm(n=2000
  , mu=rep(0,each=i)
  , Sigma=Vphy
)

@@

dd2 <- (dd
  %>% rowwise()
  %>% mutate(LR = ll-null
             # , LRT = max(as.numeric(LR > log(c)),0.5)
             , LRT = abs(LR) > c/2
             )
  # %>% arrange(LR)
)

@@

c <- qchisq(0.95,df=3-2)

dd2 <- (dd
  %>% mutate(LR = ll-null # line log ratio
       , sumLL = cumsum(ll)
       , sumnull = cumsum(null)
  )
  %>% rowwise()
  %>% mutate(LRT = abs(LR) > c/2
      , sumLR = sumLL - sumnull
      , sumLRT = sumLR > c/2
      )
  # %>% arrange(LR)
)

ddsort <- (dd
  %>% mutate(LR = ll-null) # line log ratio
  %>% arrange(LR)
  %>% mutate(sumLL = cumsum(ll)
       , sumnull = cumsum(null)
  )
  %>% rowwise()
  %>% mutate(LRT = LR > c/2
      , sumLR = sumLL - sumnull
      , sumLRT = sumLR > c/2
      )
  # %>% arrange(LR)
)
